\textit{Supplement} of ``Diagnosing the sensitivity of grounding line flux to changes in sub-ice shelf melting''

### 2019-07-24 draft_v2/supplement.tex
Updates from Overleaf

%\documentclass[10pt,a4paper]{article}
%\usepackage[latin1]{inputenc}
%\usepackage{amsmath}
%\usepackage{amsfonts}
%\usepackage{amssymb}
%\usepackage{graphicx}

% igs2ejournalguide.tex
% v4.00 3-sept-2015

\NeedsTeXFormat{LaTeX2e}

% check that the math fits the two-column format:
% \documentclass[twocolumn]{igs}

% but use this version when submitting your article:
\documentclass[review,oneside]{igs}

% other options are available
%   authors printing on US letter size are advised 
%   to use the slightly shorter [letterpaper] option
% SINGLE COLUMN
%   \documentclass{igs}              
% SINGLE COLUMN, FEWER LINES/PAGE
%   \documentclass[letterpaper]{igs} 
% DOUBLE COLUMN, FEWER LINES/PAGE
%   \documentclass[twocolumn,letterpaper]{igs}

\usepackage{igsnatbib}
\usepackage{appendix}

% check if we are compiling under latex or pdflatex
\ifx\pdftexversion\undefined
\usepackage[dvips]{graphicx}
\else
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\epstopdfsetup{suffix=}
\fi

@@

\begin{document}
\title{Diagnosising the sensitivity of grounding line flux to changes in sub-ice shelf melting}

\author[Zhang and others]{Tong ZHANG,$^1$ Stephen PRICE,$^1$ Matthew HOFFMAN,$^1$ Mauro PEREGO,$^2$ Xylar ASAY-DAVIS$^1$}
\affiliation{%
  $^1$Fluid Dynamics and Solid Mechanics Group, Los Alamos National Laboratory, New Mexico, United States, 87545, USA\\
  $^2$Center for Computing Research, Sandia National Laboratories, P.O. Box 5800, MS 1320, Albuquerque, NM 87185, USA.\\
  Correspondence: Tong Zhang 
  $<$tzhang@lanl.gov$>$}

\abstract{Motivated by previous work using ice flow models to quantify the impacts of ice shelf buttressing on grounding line flux \citep[e.g.,][]{furst2016,reese2018}, we look for an improved physical understanding for how ice dynamics link ice thickness perturbations, via changes in sub-ice shelf melting, to changes in ice shelf buttressing and grounding line flux. More specifically, we seek to define one or more ``metrics'' that are 1) readily calculated from standard ice sheet model outputs and 2) informative with respect to diagnosing the sensitivity of grounding line flux to changes in ice thickness at specific locations on an ice shelf. By studying the ice dynamics for both idealized (MISMIP+) and realistic (Larsen C) ice shelves, we find a strongly direction-dependent buttressing number that links local changes in ice shelf thickness and dynamics with changes in the integrated grounding line flux. This buttressing metric, defined using the first principal stress, is better overall for quantifying changes in grounding line flux than a similar metric defined using the second principal stress or stress along the flow direction. This correlation is possibly controlled by the relative relationship of geometric/dynamic features between the perturbation point and the grounding line, indicating a dynamic (time evolving) sensitivity field of grounding line flux change to basal melt. Unfortunately, this metric only shows a robust relationship with the integrated grounding line flux for regions near the center of an ice shelf; for points too near the grounding line or the calving front, no clear relationship exists between any of the readily calculable metrics explored here and changes in grounding line flux. This motivates our exploration of an adjoint-based method for defining grounding line flux sensitivity to local changes in ice shelf geometry. When using the same idealized case, we demonstrate that this adjoint-based method is equivalent to the sensitivity analysis of \citep{reese2018} while requiring only a single model adjoint solve. We conclude that adjoint-based methods provide a powerful and efficient tool for determining the grounding line flux sensitivity to changes in sub-ice shelf melting.}

\maketitle

\section{Introduction}

Marine ice sheets like West Antarctica (and to a lesser extent, portions of East Antarctica) are grounded below sea level and their bedrock would remain so even after full isostatic rebound \citep{barletta2018}. This and the fact that ice sheets generally thicken inland lead to a geometric configuration that are probe to instability; a small increase in flux at the grounding line thins the ice there, leading to floatation, a retreat of the grounding line into deeper water, further increases in flux (due to still thicker ice), and further thinning and grounding line retreat. This theoretical ``marine ice sheet instability'' mechanism \citep{mercer1978, schoof2007} is supported by idealized \citep[e.g.,][]{schoof2012, asay2016} and realistic \citep[e.g.,][]{Cornford2015,royston2016} ice sheet modeling experiments and some studies \citep{joughin2014,rignot2014} argue that such an instability is currently under way along outlet glaciers of Antarctica's Amundsen Sea Embayment (ASE). The relevant perturbation for grounding line retreat in the ASE is thought to be intrusions of relatively warm, intermediate-depth ocean waters onto the continental shelves, which have reduced the thickness and extent of marginal ice shelves via increased submarine melting \citep[e.g.,][]{JenkinsEtAl2016}. These reductions are critical because fringing ice shelves restrict the flux of ice across their grounding lines farther upstream---the so-called ``buttressing'' affect of ice shelves \citep{gudmundsson2012, gudmundsson2013, derydt2015}---which makes them a critical control on the rate of ice flux from Antarctica to the ocean.

%It is therefore a big concern at present if the marine-type West Antarctica Ice Sheet (WAIS) will accelerately retreat and cause a significant sea level rise in the near future \citep{joughin2011}. As ice shelves can supply additional resisting forces (buttressing) to ice flow from upstream and thus decrease the ice flux across grounding line (GL), they are critical for stabilizing marine ice sheets from further retreating \citep{gudmundsson2013}.

%Warm ocean water is getting more evident of being responsible of increased melt at the base of ice shelves in Antarctica \citep{jenkins2018,rintoul2016}. The basal melting decreases ice thickness and reduce the buttressing ability of ice shelf \citep{gagliardini2010}, and can therefore potentially destabilize marine ice sheets. Satellite altimetry measurements have proved this mechanism that ice shelf thickness decrease links glacier acceleration in Antarctica \citep{pritchard2012}. Recent studies also suggested that the thinning of ice shelfves also contributes to the ice dynamic changes of upstream grounded ice sheets \citep{minchew2018}, indicating the importance of ice shelf-ocean coupling in the whole ice sheet dynamics. 
On ice shelves, gradients in hydrostatic pressure are balanced primarily by the extensional flow of ice towards the calving front \citep{hutter1983, morland1987, schoof2007} and, in theory, an ice shelf in one horizontal dimension ($x$-$z$) provides no buttressing \citep{schoof2007,gudmundsson2013}. For realistic, three-dimensional ice shelves however,  buttressing results from three main sources: 1) compressive ice flow, 2) lateral shear, and 3) ``hoop'' stress \citep{pegler2012}. Both compressive and lateral shear stresses can supply backward resistance to extensional ice shelf flow through along-flow and across-flow stress gradients, while hoop stress is a transverse stress arising from azimuthal extension in regions of diverging flow \citep{wearing2016}.  Due to the complex geometries, kinematics, and dynamics of real ice shelves, an understanding of the specific processes and locations that control ice shelf buttressing is far from straightforward.

Several recent studies apply whole-Antarctic ice sheet models, optimized to present-day observations, towards improving our understanding for how Antarctic ice shelves limit flux across the grounding line (and by extension how they impact ice dynamics farther inland). \citet{furst2016} calculated the buttressing across Antarctica ice shelves along two major directions (aligned with the ice flow and with the second principal stress) and evaluated their impacts on upstream ice dynamics to identify regions of the ice shelves that are dynamically ``passive''; in these regions increased submarine melting, or even complete removal of the ice should not significantly alter local or regional ice dynamics or the flux of from ice upstream. \citet{reese2018} used perturbation experiments to link small, localized decreases in ice shelf thickness to changes in integrated grounding line flux (GLF), thereby providing a map of GLF sensitivity to local increases in submarine melt rates. \citet{Goldberg2019} assessed the consistency of satellite retrieved and modeled melt rates, found the importance of bathymetry in influencing the ice sheet responses predicted by models and also investigated the adjoint sensitivity of volume above floatation to melt rates for Dotson and Crosson ice shelves, Antarctica.

%SP: After writing this sentence, it occurs to me that something else that we could follow up on at some point is if / how broader patterns of ice thickness change might complicate the interpretation of these simple perturbation experiments. E.g., you can imagine that 1) multiple perturbations in diff. areas of the ice shelf might conflict with one another and have a more complex impact on g.l. flux (e.g., simultaneous perturbations in multiple areas that show both g.l. flux increases and decreases w/ ice thickness perturbations), 2) similar scale perturbations but applied over wider areas (e.g., over an area that is representative of the actual length scale of submarine melt patterns). Maybe this would be implicit in any 2nd or later study where we perturbed ice shelves based on the sensitivity maps we derive from the adjoint method.

%TZ: Yes. This is what I've been thinking of for a while. What I am not quite sure is how we design such complex (but also follow some simplied patterns) perturbation conflicts? Is there anything we can borrow from the ocean modeling results (e.g. ALCC)?

\textbf{Steve to update this paragraph to better reflect what we actually do below: 1) connect the studies of Furst and Reese by trying to understand if / how local evaluations of ice shelf buttressing do or do not reflect the actual potential for purturbations at those same locations to impact the flux across the GL (since that is, presumably, why we care about buttressing in the first place -- noting that what we'd like is some easily calculated 'metric' to identify these important locations for us); 2) assess the ability of analagous, relatively easy to calculate, local buttressing metrics to indicate GLF sensitivity to local perturbations in ice shelf geometry; 3) point out where these metrics are useful and where (and why) they are not; 4) propose and demonstrate alternate, adjoint-based method that has advantages over previous methods.}
Motivated by these studies, we build on and extend the methods and analysis of \citet{furst2016} and \citet{reese2018} in order to make progress towards answering the following questions: 
(1) How do local and regional changes in ice shelf geometry affect distal changes in GLF? 
%(2) Can local or regional ice shelf dynamics explain GLF sensitivity to local or regional changes in ice shelf thickness? 
(2) Can we derive and define new analyses for understanding how observed or modeled spatial patterns in submarine melting influence GLF and, by extension, project how changes in submarine melt pattern and magnitude will impact GLF in the future?

% OLDER: In this study we aim to further improve the understanding of how changes in submarine melting impact grounding line flux, and by inference, marine ice sheet stability. Using both idealized  (MISMIP+; Fig \ref{mismip_larsenc}a) and realistic (Larsen C; Fig \ref{mismip_larsenc}b) ice shelf geometries, we conduct model perturbation experiments similar to those of \citet{reese2018}. Our goal is to better understand the ice dynamical processes linking a local perturbation in ice shelf thickness (via submarine melt changes) to a change in grounding line flux. \textit{SP: I'm not sure if we actually do contribute much to this understanding though, so we might need to refocus this depending on what we actually include in the final version of the paper.} In addition, we explore diagnostics for quantifying grounding-line-flux sensitivity as a function of changes in submarine melt with an eye towards defining quantitative sensitivity ``metrics'' that are readily calculated from standard ice sheet model outputs.

Below, we first provide a brief description of the ice sheet model used in our study. We follow with a description of the model experiments. We then demonstrate and discuss the experimental results, including defining metrics for quantifying GLF sensitivity to changes in submarine melt, explaining their origins, and exploring their applicability and limitations.  We conclude by proposing and demonstrating an adjoint-based calculation that provides a sensitivity map analogous to the perturbation experiments from \citet{reese2018} but at the cost of a single adjoint solve in the model.

%\textbf{TZ: the Goldberg 2019 paper has already mentioned the adjoint method...Not sure yet how to revise it to avoid conflicts with their conclusions. SP: Agreed - that's why I thought we needed to mention it. But even if their application is similar, it's still fair game for us to introduce it here (we are pointing out the connection between that and Reese et al., which they do not do) as long as we acknowledge their work.}

\begin{figure}
\centering
\includegraphics[width=1\linewidth]{figs/mismip+_larsenC.pdf}
\caption{Plan view of steady-state surface speed for a) MISMIP+ and b) Larsen C Ice Shelf. The black curves show the grounding lines.}
\label{mismip+_larsenC}
\end{figure}

\section{Model description}
We use the MPAS-Albany Land Ice model \citep[MALI;][]{hoffman2018}, which solves the three-dimensional, first-order approximation to the Stokes momentum balance for ice flow\footnote{See \citet{schoof2013} for a full description of the Stokes momentum balance for ice flow and its lower-order approximations.}. Using the notation  of \citet{perego2012} and \citet{tezaur2015a} this can be expressed as, 
\begin{equation} \label{eq:foStokes}
\left\{
\begin{array}{rcl} -\nabla \cdot (2 \mu_e \dot{\boldsymbol{\epsilon}}_1) + \rho_{i} g
\frac{\partial s}{\partial x}&=&0, \\
-\nabla \cdot (2 \mu_e \dot{\boldsymbol{\epsilon}}_2) +\rho_{i} g
\frac{\partial s}{\partial y} &=& 0, \\
\end{array}\right.
\end{equation}
where $x$ and $y$ are the horizontal coordinate vectors in a Cartesian reference frame, $s(x,y)$ is the ice surface elevation, $\rho_{i}$ represents the ice density, $g$ the acceleration due to gravity, and $\dot{\boldsymbol{\epsilon}}_{1,2}$ are the two dimensional strain rate vectors given by
\begin{equation}
\dot{\boldsymbol{\epsilon}}_1 = \left(\begin{array}{ccc}
2\dot{\epsilon}_{xx} + \dot{\epsilon}_{yy}, &\dot{\epsilon}_{xy},&
\dot{\epsilon}_{xz}\end{array}\right)^T,
\end{equation}
and
\begin{equation}
\dot{\boldsymbol{\epsilon}}_2 = \left(
\begin{array}{ccc}\dot{\epsilon}_{xy}, &
\dot{\epsilon}_{xx} + 2\dot{\epsilon}_{yy}, &\dot{\epsilon}_{yz}
\end{array}\right)^T.
\end{equation}
The ``effective'' ice viscosity, $ \mu_e$ in  (\ref{eq:foStokes}), is given by 
\begin{equation}
\label{eq:effvisc}
    \mu_{e}~=~\gamma A^{-\frac{1}{n}}\dot{\epsilon}_{e}^{\frac{1-n}{n}},
\end{equation}
where $\gamma$ is an ice stiffness factor, $A$ is a temperature-dependent rate factor, $n=3$ is the power-law exponent, and the effective strain rate, $\dot{\epsilon}_{e}$, is defined as
\begin{equation} \label{eq:effstrain}
\dot{\epsilon}_e \equiv \left( \dot{\epsilon}_{xx}^2 +
\dot{\epsilon}_{yy}^2 + \dot{\epsilon}_{xx} \dot{\epsilon}_{yy} +
\dot{\epsilon}_{xy}^2 + \dot{\epsilon}_{xz}^2 +
\dot{\epsilon}_{yz}^2 \right)^\frac{1}{2}.
\end{equation}
Gradients in the horizontal velocity components, $u$ and $v$, contribute to the individual strain rate terms in (\ref{eq:effstrain}) and are given by
\begin{eqnarray} \label{eq:epsilonij}
\dot{\epsilon}_{xx} &=& \frac{\partial u}{\partial x}, \hspace{0.5cm} 
\dot{\epsilon}_{yy} = \frac{\partial v}{\partial y}, \hspace{0.5cm}
\dot{\epsilon}_{xy} = \frac{1}{2}\left(\frac{\partial u}{\partial y} + \frac{\partial v}{\partial x} \right), \nonumber\\ 
\dot{\epsilon}_{xz} &=& \frac{1}{2} \frac{\partial u}{\partial z}, \textrm{ and } \hspace{0.15cm} 
\dot{\epsilon}_{yz} = \frac{1}{2} \frac{\partial v}{\partial z}.
\end{eqnarray}

A stress free upper surface is enforced through 
%\begin{equation} \label{eq:stressFreeBC}
%\dot{\boldsymbol{\epsilon}}_1 \cdot \mathbf{n} = \dot{\boldsymbol{\epsilon}}_2 \cdot \mathbf{n} = 0, 
%\end{equation}
\begin{equation} \label{eq:stressFreeBC}
\boldsymbol{\sigma} \cdot \mathbf{n} = 0, 
\end{equation}
where $\mathbf{n}$ is the outward pointing normal vector at the ice sheet upper surface, $z=s(x,y)$.
The lower surface is allowed to slide according to the continuity of basal tractions,
%\begin{equation} \label{eq:basalbc}
%\begin{array}{ll}
%2\mu_e \dot{\boldsymbol{\epsilon}}_1 \cdot \mathbf{n} + \beta u = 0, \hspace{0.2cm} 2 \mu \dot{\boldsymbol{\epsilon}}_2 \cdot \mathbf{n} + \beta v = 0, \\
%\end{array}
%\end{equation}
\begin{equation} \label{eq:basalbc}
\boldsymbol{\sigma} \cdot \mathbf{n} + \beta \boldsymbol{u} = 0,
\end{equation}
where $\beta$ is a spatially variable friction coefficient, $\boldsymbol{\sigma}$ is the stress tensor and $\boldsymbol{u}$ is the two-dimensional velocity vector ($u$, $v$). On lateral boundaries in contact with the ocean, the portion of the boundary above sea level is stress free while the portion below sea level feels the ocean hydrostatic pressure according to
%\begin{equation}\label{eq:oceanbc}
%\begin{array}{ll}
%2 \mu_e \left(\dot{\boldsymbol{\epsilon}}_1 \cdot \mathbf{n},\, \dot{\boldsymbol{\epsilon}}_2 \cdot \mathbf{n},\, 0\right)^T -\rho_i g (s-z)\mathbf{n} = \rho_o g \max(z,0) \mathbf{n},
%\end{array}
%\end{equation}
\begin{equation}\label{eq:oceanbc}
\boldsymbol{\sigma} \cdot \mathbf{n} = \rho_i g (s-z)\mathbf{n},
\end{equation}
%where $\rho_o$ represents the density of ocean water and $xx\mathbf{n}$ the outward pointing normal vector to the lateral boundary (i.e., parallel to the $(x,y)$ plane).
where $\mathbf{n}$ the outward pointing normal vector to the lateral boundary (i.e., parallel to the $(x,y)$ plane).
%\textbf{TZ: I change this section a bit to avoid it looking identical to Matt's MALI paper.}

A more complete description of the MALI model, including the implementations for mass and energy conservation, can be found in \citet{hoffman2018}. Additional details about the Albany momentum balance solver can be found in \citet{tezaur2015a,tezaur2015b}.

Here, we apply MALI to experiments on both idealized and realistic marine-ice sheet geometries. For our idealized domain and model state, we start from the equilibrium initial conditions for the MISMIP+ experiments, as described in \citet{asay2016}. The model mesh is spatially uniform at 2 km resolution. For our realistic domain, we use Antarctica's Larsen C Ice Shelf and its upstream catchment area. The model state is based on the optimization of the ice stiffness ($\gamma$ in Eqn (\ref{eq:effvisc})) and basal friction ($\beta$ in Eqn (\ref{eq:basalbc})) coefficients in order to provide a best match between modeled and observed present-day velocities \citep{rignot2014} using adjoint-based methods discussed in \citet{perego2014} and \citet{hoffman2018}. The domain geometry is based on Bedmap2 \citep{fretwell2013} and ice temperatures, which are held fixed for this study, are based on \citet{liefferinge2013}. Mesh resolution on the ice shelf is between 2 and 6 km and coarsens to 20 km in the ice sheet interior. Following optimization to present-day velocities, the model is relaxed using a 100-year forward run, providing the initial condition from which the Larsen C experiments are conducted (as discussed below). Both the MISMIP+ and Larsen C experiments use 10 vertical layers that are finest near the bed and coarsen towards the surface. The grounding line position is determined from hydrostatic equilibrium. A sub-element parameterization method is used to define basal friction coefficient values at the grounding line \citep{seroussi2014} .

% For the idealized MISMIP+ experiments, we use its  steady-state geometry following the same model configurations as in \citet{asay2016}. The mesh is spatially uniform (2 km). Both MISMIP+ and Larsen C experiments use 10 vertical layers that are finest near the bed (4\% of total thickness) and coarsen towards the surface (23\% of total thickness).

\section{Perturbation experiments}

To explore the sensitivity of changes in GLF to small, localized changes in ice shelf thickness, we conduct a number of perturbation experiments analogous to those of \citet{reese2018}. Using diagnostic (instantaneous) model solutions, we first study the instantaneous response of GLF for the idealized geometry and initial state provided by the MISMIP+ experiment \citep{asay2016}. We then conduct a similar study for Antarctica's Larsen C Ice Shelf using a realistic configuration and initial state.

Our experiments are conducted in a manner similar to those of \citet{reese2018}. We perturb the coupled ice sheet-shelf system by decreasing the ice thickness uniformly by 1 m over square grid ``boxes'' covering the base of the ice shelves, after which we examine the instantaneous impact on kinematics and dynamics (discussed further below). For MISMIP+, we use a uniform, hexagon-type, around 2-km mesh (for simplicity we use the term 2-km in the following). For the Larsen C Ice Shelf, horizontal mesh resolution is spatially variable and we assign each grid cell to fall within one and only one box based on its location. For MISMIP+, we perturb the thickness of the cell (hexagon) in the mesh. For Larsen C, we only use 20$\times$20 km square boxes \citep[the same resolution as in][]{reese2018}. Lastly, for the MISMIP+ 2-km experiments we note that, in order to save on computing costs, we only perturb the region of the ice shelf for which $x<530$ km (the area over which the ice shelf is laterally buttressed).  We also perturb only half of the ice shelf, $y>40$ km, due to the symmetry about the center line, though we analyze the response to these perturbations over the entire model domain.

Similarly to \citet{reese2018}, we define a GLF ``response number'',
\begin{equation}
N_r = \left(\frac{R}{P}\right)^k,
\end{equation}
where $R$ is the ice flux change integrated along the entire grounding line, $P$ is the mass change associated with a single grid-box perturbation and $k$ is a power-law index that allows for the possibility of a nonlinear relationship between ice shelf buttressing and the change in GLF \citep[see also][]{schoof2007}. Here, we use $k=1/n$ with $n=3$.

Despite the existence of many possible factors linking GLF to ice shelf properties (ice flow direction, horizontal gradients in ice shelf geometry, stress fields, strain-rate fields, perturbation locations, etc.), here we mainly examine model stress fields and the distance between perturbation points and the GL. This is because, as we show below, these factors correlate closely with the sensitivity of changes in GLF to the imposed ice shelf perturbations. \textbf{SP: Maybe we can skip the previous sentences and just skip to some form of the next one? We can motivate this simply by saying that the reason we use a buttressing number like Furst et al is that it's a popular/standard and easy way to connect local ice shelf perturbations to the changes in GLF from the Reese experiments.} To incorporate the local stress field and its buttressing capacity into our analysis, we also calculate a ``buttressing number'', $N_b$, analogous to that from \citet{furst2016},
\begin{equation}
N_b\left(\mathbf{n}\right)=1-\frac{\mathbf{n}\cdot\mathbf{\sigma}\cdot\mathbf{n}}{N_0},
\label{butN}
\end{equation}
where $N_0$ is the vertically integrated ocean pressure ($N_0=\frac{1}{2}\left(1-{\rho_i}/{\rho_w}\right)gH$) and $\rho_i$ (910\,kg\,m$^{-3}$) and $\rho_w$ (1028\,kg\,m$^{-3}$) are the density of ice and ocean water, respectively. $\sigma_{nn}=\mathbf{n}\cdot\mathbf{\sigma}\cdot\mathbf{n}$ is the normal stress along a specified horizontal direction given by the normal vector $\mathbf{n}$ (see details in below).

%\begin{table*}
%\centering
%\caption{Stress components used in the buttressing (Eqn \ref{butN})}
%\label{stressDef}
%\begin{tabular}{|c|c|c|c|}
%\hline
%$\sigma_{obj}$ & def. & $\sigma_{obj}$ & def. \\
%\hline
%$\sigma_f$ & ${\bf{n}}^T_f\cdot{\bf{\sigma}}\cdot{\bf{n}}_f$ & $\sigma_{af}$ & ${\bf{n}}^T_{af}\cdot{\bf{\sigma}}\cdot{\bf{n}}_{af}$ \\ 
%\hline 
%$\sigma_{x}$ & ${\bf{n}}^T_x\cdot{\bf{\sigma}}\cdot{\bf{n}}_x$ & $\sigma_{y}$ & ${\bf{n}}^T_y\cdot{\bf{\sigma}}\cdot{\bf{n}}_y$ \\ 
%\hline
%$\sigma_{p1}$ & $\frac{\sigma_{x}+\sigma_{y}}{2}+\sqrt{\left(\frac{\sigma_{x}-\sigma_{y}}{2}\right)^2+\sigma_{xy}^2}$ & $\sigma_{p2}$ & $\frac{\sigma_{x}+\sigma_{y}}{2}-\sqrt{\left(\frac{\sigma_{x}-\sigma_{y}}{2}\right)^2+\sigma_{xy}^2}$ \\ 
%\hline 
%$\sigma_s$ & ${\bf{n}}^T_f\cdot{\bf{\sigma}}\cdot{\bf{n}}_{af}$ & $\sigma_h$ & $\eta H\frac{\partial}{\partial r}\left(\frac{u}{r}\right)$ \\ 
%\hline
%
%\end{tabular}
%\end{table*}

\section{Results and Discussion}

%\subsection{Linear relationship between buttressing ($N_b$) and GL flux responses ($N_r$)}
\subsection{Correlation between buttressing and changes in GLF}

A decrease in ice shelf buttressing tends to lead to an increase in GLF \citep[e.g.,][]{gagliardini2010} and intuitively we expect that the GLF should be relatively more sensitive to melt changes that occur in regions of relatively larger buttressing. Here, we aim to better understand and quantify the relationship between the local ice shelf buttressing ``strength'' (given by $N_b$) and changes in GLF (given by $N_r$). Can the change in GLF for a given melt perturbation be quantified (in the predictive sense) as a function of the buttressing number? In Fig.~\ref{mismip_Nb_GLF_regression}, we show correlations between $N_b$ and $N_r$ for perturbations applied to the MISMIP+ domain when $N_b$ from (\ref{butN}) is defined based on the direction of the first principal stress ($\sigma_{p1}$) \footnote{We expand on our reasons for choosing $\sigma_{p1}$ below.}. In Fig.~\ref{mismip_Nb_GLF_regression}b, we include all perturbed ice shelf locations in the comparison and find relatively weak $N_b$:$N_r$ correlations. In Fig.~\ref{mismip_Nb_GLF_regression}c, we remove from the comparison points that are 1) weakly buttressed ($x>480$\,km, where the ice shelf becomes unconfined) and 2) within 10 km of the GL, and a much stronger $N_b$:$N_r$ correlation emerges. In Fig.~\ref{mismip_Nb_GLF_regression}d, we plot the $N_b$:$N_r$ correlation for locations where $x>480$\,km and the magnitude of the thickness gradient $\left|\nabla H\right|<7\times 10^{-3}$ (a similar plot with the buttressing number defined based on the direction of the second principal stress is shown in Fig.~\ref{sigma_p2_example}). The strong similarities between Figures \ref{mismip_Nb_GLF_regression}c and \ref{mismip_Nb_GLF_regression}d suggest that the seemingly \textit{ad hoc} ``distance from the grounding line'' constraint is important for removing areas of complex geometry (and hence complex dynamics) from the $N_b$:$N_r$ comparisons. We discuss possible reasons for this further below.

% in (\ref{butN})) in the MISMIP+ test case (Fig \ref{mismip_Nb_GLF_regression}a), we find relatively weak $N_b$-$N_r$ correlations when we include the results from all perturbation experiments (Fig \ref{mismip_Nb_GLF_regression}b). However, if we do not consider the perturbation points that are weakly buttressed ($x>480$ km; near the open shelf region) and are close to GL (the minimum distances between GL and perturbation points are greater than 5 km), we can clearly see a strong linear $N_b$-$N_r$ regression (Fig \ref{mismip_Nb_GLF_regression}c). A similar strong linearity can also be found if we apply a different filtering approach using the ice thickness gradient field (Fig \ref{mismip_Nb_GLF_regression}a), i.e., we keep the perturbation points that have thickness gradients smaller than 0.0007 (Fig \ref{mismip_Nb_GLF_regression}d). This nonlinearity feature is likely caused by the intensive shearing and complex ice flow mechanics near GL (a strong transition zone from relatively slow grounded ice flow to fast extensively floating ice flow). We can again see this non-linearity region in our adjoint sensitivity experiments (Fig.~\ref{fig9} in section ``Adjoint sensitivity'').

\begin{figure}
\centering
\includegraphics[width=1\linewidth]{figs/mismip_Nb_GLF_regression.pdf}
    \caption{(a) MISMIP+ steady-state geometry. Color represents the magnitude of the ice thickness gradient and the white line represents the GL. (b) $N_b$:$N_r$ correlation for all perturbation points. (c) $N_b$:$N_r$ correlation for perturbation points within the confined region of the shelf ($x>480$) and $>$10 km from the GL; (d) $N_b$:$N_r$ correlation for perturbation points within the confined region of the shelf and with a thickness gradient magnitude $\left|\nabla H\right|<7$x$10^{-3}$. In (c) and (d), the orange dots in the insets represent the GL grid cells, and the blue dots the perturbation grid cells. \textbf{SP: Matt suggested expanding this to include 2 more panels using the points in (d) -- the flow direction and the p2 direction (the goal being to show that the correlations are pretty bad). Then we could also add another panel along the bottom which is the 'continuous' version of these (i.e., the red curve in Fig. 3a).}}
\label{mismip_Nb_GLF_regression}
\end{figure}

%\subsection{The buttressing directions}
%\subsection{Correlation dependence on buttressing direction}
\subsection{Directional dependence of buttressing}

The buttressing number, $N_b$, is computed using the normal stress ($\sigma_{nn}$) projected onto a specific direction, see Equation (\ref{butN}). Therefore, the buttressing number at any perturbation point varies depending on the chosen direction. \citet{furst2016} calculated $N_b$ along two directions, the ice flow direction ($\mathbf{n}_f$) and the direction corresponding to the second principal stress ($\sigma_{p2}$) ($\mathbf{n}_{p2}$). They found that the latter---the direction corresponding to the maximum compressive stress (or the least extensional stress)---best identified ``passive'' regions, where the ice shelf provides little or no buttressing. In Fig.~\ref{mismip_r2_all_direction}, we plot the correlation coefficients ($r^2$) between $N_r$ and $N_b$ where the direction corresponding to $\sigma_{nn}$ varies continuously between 0 and 180$^\circ$ (as a function of the angle $\Delta \phi$) relative to the direction corresponding to the first principle stress ($\sigma_{p1}$) ($\mathbf{n}_{p1}$). We find the largest correlation coefficient ($r^2>0.9$) when $N_b$ is aligned with $\mathbf{n}_{p1}$ ($\Delta \phi=0^\circ$) and the smallest correlation coefficient ($r^2<0.4$) when $N_b$ is aligned with $\mathbf{n}_{p2}$ ($\Delta \phi=90^\circ$). We can also see the corresponding variations in buttressing number $N_b$ in Fig.~\ref{Nb_Deltaphi}, where the maximum buttressing number is calculated from $N_r(\mathbf{n}_{p2}$) according to Equation (\ref{butN}).
% for different directions with respect to the first principal stress and flow direction, respectively (Fig \ref{mismip_r2_all_direction}). Differently, we find that $N_b$ along the $\sigma_{p1}$ direction shows the best regression performance, whereas the $\sigma_{p2}$ appears to show the weakest correlations to $N_r$. This can be further testified by looking at the angle differences between $\sigma_{p1}$ and flow directions ($\Delta \phi = \phi_{flow}-\phi_{\sigma_{p1}}$).

>>>>

\documentclass[tc, manuscript]{copernicus}

%\usepackage{xr}
%\usepackage{hyperref}
%\usepackage{appendix}

% check if we are compiling under latex or pdflatex
%\ifx\pdftexversion\undefined
%\usepackage[dvips]{graphicx}
%\else
%\usepackage[pdftex]{graphicx}
%\usepackage{epstopdf}
%\epstopdfsetup{suffix=}
%\fi

@@

\begin{document}
\title{\textit{Supplement} of ``Diagnosing the sensitivity of grounding line flux to changes in sub-ice shelf melting''}

\Author[1]{Tong}{Zhang}
\Author[1]{Stephen}{Price}
\Author[1]{Matthew}{Hoffman}
\Author[2]{Mauro}{Perego}
\Author[1]{Xylar}{Asay-Davis}

\affil[1]{Fluid Dynamics and Solid Mechanics Group, Los Alamos National Laboratory, Los Alamos, NM, 87545, USA}
\affil[2]{Center for Computing Research, Sandia National Laboratories, Albuquerque, NM, 87185, USA}

\runningtitle{Diagnosising the sensitivity of grounding line flux to sub-ice shelf melting}
\runningauthor{T.~Zhang et al.}
\correspondence{T.~Zhang (tzhang@lanl.gov)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

\firstpage{1}
\maketitle

\section{Appendix materials}
%\renewcommand\thefigure{\thesection.\arabic{figure}}  
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}
\centering
\includegraphics[width=1\linewidth]{./figs/mismip_Nb_GLF_regression_all_points.pdf}
    \caption{(a) Blue dots are all perturbation points for $N_r\left(N_b\right)$ linear regression analysis and the black dots represent grid cells along the grounding line. (b) Modeled $N_r$ from perturbation experiments versus fitted $N_r$ as a function of $N_b$ that is calculated along $\textbf{n}_{p1}$; (c) Modeled $N_r$ from perturbation experiments versus fitted $N_r$ as a function of $N_b$ that is calculated along $\textbf{n}_{p2}$; (d) Modeled $N_r$ from perturbation experiments versus fitted $N_r$ as a function of $N_b$ that is calculated along $\textbf{n}_{f}$. $r$ is the correlation coefficient. \textbf{Tong: Do we need to include the information of significant level along with correlation coefficient?}}
\label{mismip_Nb_GLF_regression_all_points}
\end{figure}

%\begin{figure}
%	\centering
%	\includegraphics[width=1\linewidth]{figs/sigma_p2_example.pdf}
%	\caption{$N_b$:$N_r$ correlation for perturbation points within the confined region of the shelf and with a thickness gradient magnitude $\left|\nabla H\right|<7$x$10^{-3}$. $N_b$ is calculated from the second principal stress $\sigma_{p2}$.}
%	\label{sigma_p2_example}
%\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{./figs/Nb_Deltaphi.pdf}
	\caption{(a) $N_b$ values rotated counterclockwise by $\Delta\phi$ degrees relative to the direction corresponding to $\sigma_{p1}$ ($\mathbf{n}_{p1}$); (b) $N_b$ values rotated counterclockwise by $\Delta\phi$ degrees relative to the direction corresponding to $\sigma_{p1}$ ($\mathbf{n}_{f}$). Note that these are mean $N_b$ values for those perturbation points shown in Fig.~\ref{mismip_Nb_GLF_regression}a. It's a twin plot for Fig.~\ref{mismip_Nb_GLF_regression}a and b. Both curves show the buttressing number is largest along the direction corresponding to $\sigma_{p2}$ ($\mathbf{n}_{p2}$), i.e., in (a) $N_b$ is the largest when $\Delta\phi\approx 90^\circ$ relative to $\mathbf{n}_{p1}$, and in (b) $N_b$ is the largest when $\Delta\phi\approx 62^\circ$ relative to $\mathbf{n}_{f}$ which is $\Delta\phi+\Delta\varphi$ (around 92--102$^\circ$) relative to $\mathbf{n}_{p1}$ ($\Delta\varphi$ is the angle difference between $\mathbf{n}_{f}$ and $\mathbf{n}_{p1}$, around 30--50$^\circ$; see Section~\ref{directional_buttressing}).}
	\label{Nb_Deltaphi}
\end{figure}
